Indel Historian: fast and accurate ancestral sequence reconstruction using weighted finite-state transducers

### 2016-07-22 doc/main.tex
writing

\title[Indel Historian: phylogenetic ancestral reconstruction]{Indel Historian: phylogenetic ancestral reconstruction}
\author[Ian Holmes]{Ian Holmes$^{\text{\sfb 1}*}$}
\address{$^{\text{\sf 1}}$Department of Bioengineering, University of California, Berkeley, 94703, USA.}

@@

Multiple sequence alignments are used for different purposes in bioinformatics,
and these different applications have led to different criteria for evaluating
the accuracy of multiple alignment tools

Homology modeling in protein structure prediction
leads to accuracy metrics that quantify the number of correctly aligned residues
compared to a benchmark dataset of structurally-informed reference alignments

Significant progress has been made optimizing these metrics

Successful techniques include using Bayesian decision theory to maximize the expected
score under the metrics \citep{NotredameEtAl2000,DoEtAl2005,SchwartzPachter2007,BradleyEtAl2009}
and fine-tuning the alignment scoring function to reflect known details of
protein sequences under selection for folding, such as the reduced likelihood of indels in
hydrophobic regions \citep{KatohEtAl2005,Edgar2004b,LarkinEtAl2007}

PRANK\citep{LoytynojaGoldman2005}, Muscle \citep{Edgar2004b}, ClustalW\citep{LarkinEtAl2007}, MAFFT\citep{KatohEtAl2005}

These have been accompanied by performance optimizations at the rate-limiting steps
of multiple alignment, including all-versus-all sequence comparison \citep{BradleyEtAl2009,Edgar2004b}

The study of sequence evolution and molecular phylogenetics gives rise to
an alternative set of criteria for measuring alignment accuracy

Previous work showed that, for the purposes of estimating
molecular evolutionary parameters---such as the rates of insertion-deletion events \citep{Westesson2012-zg},
the ratio between synonymous and nonsynonymous substitution rates \citep{Redelings2014},
or other evolutionary analyses \citep{LoytynojaGoldman2008}---it is best to use an explicit probabilistic model of the sequence evolution process
and to treat sequence alignment rigorously as a missing data estimation problem
under this model

The ProtPal program is one such approach which uses the theory of
weighted finite-state transducers
These are automata which can be multiplied like substitution matrices
and so can be thought of as a strict generalization of Felsenstein's likelihood-phylogenetic
approach to multiple sequence alignment
(in fact, the use of transducers makes Felsenstein's pruning algorithm formally
equivalent to a special case of Sankoff's multiple alignment algorithm)

A previously published benchmark of alignment methods on simulated sequence data
showed that most alignment algorithms introduce systematic biases into the
estimation of indel rates, and that these biases are typically substantially worse
at higher indel rates \citep{Westesson2012-zg}.

ProtPal was found to be the most accurate algorithm for indel rate estimation,
followed by PRANK and then MUSCLE.

However, the implementation of ProtPal published with that benchmark was too slow
for practical use, and no benchmark of ProtPal on structural alignment benchmarks was available

Here, we present a clean implementation of the algorithm underlying ProtPal
in a new tool, Indel Historian
The tool is targeted to evolutionary applications such
as ancestral sequence reconstruction and molecular phylogenetics

We also report an assessment of the structural alignment accuracy of IndelHistorian on the benchmark datasets BAliBase 3.0 \citep{ThompsonEtAl2005}
and PREFAB 4.0 \citep{Edgar2010}

>>>>

\title[Indel Historian: phylogenetic ancestral reconstruction]{Indel Historian: fast and accurate ancestral sequence reconstruction using weighted finite-state transducers}
\author[Ian Holmes]{Ian Holmes$^{\text{\sfb 1}*}$}
\address{$^{\text{\sf 1}}$Department of Bioengineering, University of California, Berkeley, 94703, USA.}

@@

Synthesis of reconstructed ancestral proteins offers a direct way to test hypotheses of evolutionary adaptation \citep{UgaldeEtAl2004,OrtlundEtAl2007,GaucherEtAl2008}
or to generate diversity in combinatorial libraries for bioengineering by directed evolution \citep{AlcolombriEtAl2011,SantiagoOrtizEtAl2015}.
The computational workflow for such reconstructions typically involves building a multiple sequence alignment and then using statistical phylogenetics to reconstruct the substitution history
\citep{Liberles2007}.
However, it remains an open question whether the multiple alignment tools that perform best on the most commonly-used benchmarks
(which generally use protein structure as a gold standard)
are also the best alignment tools for performing ancestral sequence reconstruction.

Multiple sequence alignments are used for several purposes in bioinformatics,
and these different applications have led to different criteria for evaluating
the accuracy of the tools.
Homology modeling in protein structure prediction
leads to accuracy metrics that quantify the number of correctly aligned residues
compared to a benchmark dataset of structurally-informed reference alignments \citep{ThompsonEtAl2005}.
Techniques that have proven successful in this context include using Bayesian decision theory to maximize the expected
score under the metrics \citep{NotredameEtAl2000,DoEtAl2005,SchwartzPachter2007,BradleyEtAl2009}
and fine-tuning the alignment scoring function to reflect known details of
protein sequences under selection for folding, such as the reduced likelihood of indels in
hydrophobic regions \citep{KatohEtAl2005,Edgar2004b,LarkinEtAl2007}.
These have been accompanied by performance optimizations of the various steps involved in multiple alignment,
the rate-limiting step generally being all-versus-all sequence comparison \citep{BradleyEtAl2009,Edgar2004b}.

The study of sequence evolution and molecular phylogenetics gives rise to
a different set of criteria for measuring alignment accuracy.
As an example, for the purposes of predicting a protein structure,
functionally equivalent sequences in a rapidly-evolving region (such as a loop)
may be considered alignable even if they have been entirely replaced by a series of indel mutations
(e.g. the sequences $X$ and $Y$ in the mutation trajectory $AXB \to AXYB \to AYB$),
but for the purposes of evolutionary analysis, one generally wants to count every indel event that has occurred.
Previous work showed that, for the purposes of estimating
molecular evolutionary parameters---such as the rates of insertion-deletion events \citep{Westesson2012-zg},
the ratio between synonymous and nonsynonymous substitution rates \citep{Redelings2014},
or other evolutionary analyses \citep{LoytynojaGoldman2008}---it is best to use an explicit probabilistic model of the sequence evolution process
and to treat sequence alignment rigorously as a missing data estimation problem
under this model.

The ProtPal program \citep{Westesson2012-zg} is one such approach which uses the theory of
weighted finite-state transducers,
automata which can be multiplied together like substitution matrices \cite{BouchardCote2013}.
This approach can be represented as a strict generalization of Felsenstein's celebrated
algorithm for computing likelihoods on phylogenetic trees \citep{Felsenstein81}.
In fact, the use of transducers makes Felsenstein's pruning algorithm formally
equivalent to a special case of Sankoff's phylogenetic multiple alignment and RNA folding algorithm \citep{Sankoff85};
the RNA folding component is included if one generalizes from finite-state transducers to pushdown automata \citep{BradleyHolmes2009}.

Our earlier benchmark of alignment methods on simulated sequence data
showed that most alignment algorithms introduce systematic biases into the
estimation of indel rates, and that these biases are typically substantially worse
at higher indel rates \citep{Westesson2012-zg}.
ProtPal was found to be the most accurate algorithm for indel rate estimation,
followed by PRANK and then MUSCLE.
However, the implementation of ProtPal published with that benchmark was too slow
for practical use.
Furthermore, no benchmark of ProtPal on structural alignment benchmarks was available,
and---although we have noted that structure prediction is a different task---it would be
useful to know just how different it is, according to the metrics.

Here, we present a clean implementation of the algorithm underlying ProtPal
in a new tool, Indel Historian.
The tool is targeted to evolutionary applications such
as ancestral sequence reconstruction and molecular phylogenetics.
We also report an assessment of the structural alignment accuracy of IndelHistorian on the benchmark datasets BAliBase 3.0 \citep{ThompsonEtAl2005}
and PREFAB 4.0 \citep{Edgar2010}.
